---
title: 'Lab 05: Fonts & Tables'
author: "Alison Hill & Steven Bedrick"
subtitle: "BMI 5/625"
always_allow_html: true
output:
  html_document:
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
    toc_float: yes
  # pdf_document:
    # toc: yes
    # toc_depth: '2'
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, comment = NA, warning = FALSE, errors = FALSE, message = FALSE, tidy = FALSE, cache = FALSE)
```

# Goals for Lab 05

Goal: Become familiar with tools for generating publication-ready tables directly in R.

We will use data from the following paper: MacFarlane, H., Gorman, K., Ingham, R., Presmanes Hill, A., Papadakis, K., Kiss, G., & van Santen, J. (2017). [Quantitative analysis of disfluency in children with autism spectrum disorder or language impairment.](http://doi.org/10.1371/journal.pone.0173936) PLoS ONE, 12(3), e0173936. 

```{r load_packages, include = FALSE}
library(tidyverse)
library(janitor)
library(knitr)
library(DT)
library(xtable)
library(broom)
library(tableone)
library(formattable)
library(pnwflights14) # devtools::install_github("ismayc/pnwflights14")
library(pixiedust)
library(kableExtra)
library(huxtable)
```

```{r}
mazes <- read_csv("data/mazes.csv") %>% 
  clean_names()  #janitor package

glimpse(mazes)

```


# TL;DR

The workhorse for making simple tables in R Markdown documents is the `knitr` package's `kable` function. This function is really versatile, but also free of fancy formatting options, for better or worse.

# `knitr::kable`

## `kable` all tables everywhere

In order to tell RMarkdown to use `kable` to format all tabular output, update the YAML header in your document and override the `df_print` option, when producing HTML output:

```{r eval = FALSE}
---
title: "My Awesome Data Vis Lab"
output:
  html_document:
    df_print: kable
---
```

You can also define the html format in the global options.
```{r eval = FALSE}
# If you don't define format here, you'll need put `format = "html"` in every kable function.
options(knitr.table.format = "html") 
# You may also wish to set this option, to handle number formatting
options(scipen = 1, digits = 2)
```


## `kable` table in a chunk

For HTML output:

```{r}
head(mazes) %>% 
  kable(format = "html")
```

Note that this is a bit rough-looking. We will use various features of `kable` to improve our output.

For starters, we can add a caption:

```{r}
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "A table produced by kable.")
```

We can also manually specify human-readable column names:

```{r}
my_maze_names <- c("Participant", "Age", "Verbal\nIQ", "Group", "Activity", "Content\nMaze", "Filler\nMaze", "Repetition", "Revision", "False\nStart", "Cued", "Not\nCued")
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "A table produced by kable.",
        col.names = my_maze_names)
```


## Styled `kable` tables in a chunk

To improve the visual layout of the table, we can use the `kableExtra` package, which provides the `kable_styling()` function:

```{r}
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "A styled kable table.",
        col.names = my_maze_names) %>% 
  kable_styling()
```

There are lots of printing options: https://haozhu233.github.io/kableExtra/awesome_table_in_html.html

```{r}
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "A non-full width zebra kable table.") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Note that by default the table will be centered; you can override this by specifying a `position`:

```{r}
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "Over here!") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

## Controlling column appearance

We can control the formatting of individual columns using the `column_spec()` function:

```{r}
head(mazes) %>% 
  kable(format = "html", digits = 2, caption = "Over here!") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(4, width="3cm", background="lightblue", border_right=TRUE)
```

## `kable` + `kableExtra` + `formattable`

`color_tile` and `color_bar` are neat extras, if used wisely and in moderation!

http://haozhu233.github.io/kableExtra/use_kableExtra_with_formattable.html

```{r}
library(formattable)
head(mazes) %>% 
  mutate(ca = color_tile("transparent", "lightpink")(ca),
         viq = color_bar("lightseagreen")(viq)) %>%
  kable("html", escape = F, caption = 'This table is colored.') %>%
  kable_styling(position = "center") %>% 
  column_spec(4, width = "3cm") 
```


## `tibble` + `kable` + `kableExtra`

Sometimes, when writing a document or preparing a report, you may have tabular information to present that is not already in a dataframe. 
For example, consider the helpful table of vectorized math operators from one of the slide decks.
I made this by first manually creating a data frame to hold the contents of the table (using the [`tibble` package's](http://tibble.tidyverse.org) `tribble()` function), and then formatting it as before.

Manually creating a quick-and-dirty dataframe with `tribble()` is simple:

```{r}
math_table <- tibble::tribble(
  ~Operator, ~Description, ~Usage, # column names
  "\\+", "addition", "x + y", # each column, for each row, separated by commas
  "\\-", "subtraction", "x - y",
  "\\*", "multiplication", "x * y",
  "/", "division", "x / y",
  "^", "raised to the power of", "x ^ y",
  "abs", "absolute value", "abs(x)",
  "%/%", "integer division", "x %/% y",
  "%%", "remainder after division", "x %% y"
)
```

Then I used this chunk to print it:

````
```{r, results = 'asis'}`r ''`
knitr::kable(math_table, format = "html", caption = "Helpful mutate functions") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```
````

```{r results = 'asis'}
knitr::kable(math_table, format = "html", caption = "Helpful mutate functions") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

# Markdown Tables

Alternatively, you may just want to type in a table in Markdown and ignore R. Four kinds of tables may be used. The first three kinds presuppose the use of a fixed-width (monospaced) font, such as Courier. The fourth method can be used with proportionally spaced fonts, as it does not require lining up columns. All of the below will render when typed *outside* of an R code chunk since these are based on `pandoc` being used to render your markdown document. Note that these should all work whether you are knitting to either html or PDF.


## Simple table

This code for a simple table:


```{r eval = FALSE}
  Right     Left     Center     Default
-------     ------ ----------   -------
     12     12        12            12
    123     123       123          123
      1     1          1             1

Table:  Demonstration of simple table syntax.
```

Produces this simple table:

  Right     Left     Center     Default
-------     ------ ----------   -------
     12     12        12            12
    123     123       123          123
      1     1          1             1

Table:  Demonstration of simple table syntax.

The headers and table rows must each fit on one line. Column alignments are determined by the position of the header text relative to the dashed line below it:3

* If the dashed line is flush with the header text on the right side but extends beyond it on the left, the column is right-aligned.
* If the dashed line is flush with the header text on the left side but extends beyond it on the right, the column is left-aligned.
* If the dashed line extends beyond the header text on both sides, the column is centered.
* If the dashed line is flush with the header text on both sides, the default alignment is used (in most cases, this will be left).
* The table must end with a blank line, or a line of dashes followed by a blank line.

The column headers may be omitted, provided a dashed line is used to end the table.

## Multi-line tables

This code for a multi-line table:

```{r eval = FALSE}
-------------------------------------------------------------
 Centered   Default           Right Left
  Header    Aligned         Aligned Aligned
----------- ------- --------------- -------------------------
   First    row                12.0 Example of a row that
                                    spans multiple lines.

  Second    row                 5.0 Here's another one. Note
                                    the blank line between
                                    rows.
-------------------------------------------------------------

Table: Here's the caption. It, too, may span
multiple lines.
```

Produces this multi-line table:

-------------------------------------------------------------
 Centered   Default           Right Left
  Header    Aligned         Aligned Aligned
----------- ------- --------------- -------------------------
   First    row                12.0 Example of a row that
                                    spans multiple lines.

  Second    row                 5.0 Here's another one. Note
                                    the blank line between
                                    rows.
-------------------------------------------------------------

Table: Here's the caption. It, too, may span
multiple lines.

## Grid tables

This code for a grid table:

```{r eval = FALSE}
: Sample grid table.

+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| Bananas       | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - tasty            |
+---------------+---------------+--------------------+
```

Produces this grid table:

: Sample grid table.

+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| Bananas       | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - tasty            |
+---------------+---------------+--------------------+

Alignments are not supported, nor are cells that span multiple columns or rows. 

Note that if you find yourself making a lot of these kinds of tables, you may want to use software like [Monodraw](https://monodraw.helftone.com) to help.

## Pipe tables

This code for a pipe table:

```{r eval = FALSE}
| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|   12  |  12  |    12   |    12  |
|  123  |  123 |   123   |   123  |
|    1  |    1 |     1   |     1  |

  : Demonstration of pipe table syntax.
```

Produces this pipe table:

| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|   12  |  12  |    12   |    12  |
|  123  |  123 |   123   |   123  |
|    1  |    1 |     1   |     1  |

  : Demonstration of pipe table syntax.

This method does not depend on using a monospaced font!


# `gt`

`gt` is a [package from RStudio](https://gt.rstudio.com/reference/cols_align.html) designed for publication-ready figures. It is my go-to tool for tables.

<img src="https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg"/>

```{r}
library(gt)
```

Let's set up a tiny little data table...

```{r}
prepped_flight_data <- flights %>% filter(dest %in% c("DEN", "DCA", "JFK", "SFO")) %>% 
  group_by(dest) %>% 
  mutate_at(vars(dep_delay), replace_na, replace=0.0) %>%  
  summarise(mean_delay=mean(dep_delay, na.rm=TRUE), median_delay=median(dep_delay, na.rm=TRUE))
glimpse(prepped_flight_data)
```
... and now make a table out of it:
```{r}
gt_formatted <- prepped_flight_data %>%  gt() %>% 
  tab_header(title="PDX Departure Delays, DCA/DEN/JFK/SFO", subtitle="(Delay in minutes)") %>% 
  fmt_number(columns=vars(mean_delay, median_delay), decimals=2) %>% 
  cols_label(dest="Destination", mean_delay="Mean", median_delay="Median") 
gt_formatted 
```

We can adjust things like text alignment after-the-fact:

```{r}
gt_formatted <- gt_formatted %>% 
  cols_align(align="center", columns=vars(dest)) %>%
  tab_style(
    cell_text(align = "right"), 
    locations = cells_title(groups = c("subtitle"))
  ) %>% 
  tab_style(
    cell_text(style = "italic"),
    locations = cells_title(groups=c("title"))
  )
gt_formatted
```

**Tip:** In addition to `cells_title()`, there are helpers to select any of the other cell types (e.g. `cells_column_labels()`, etc.).

Now we can turn it into whatever we need! As Latex:

```{r eval=FALSE}
gt_formatted %>% gtsave("my_table.tex")

```

As RTF, to put in Word:

```{r=FALSE}
gt_formatted %>% gtsave("my_table.rtf")
```

# Descriptive Stats tables

## `tableone`

`tableone` is for generating "Table 1" in your paper. You know, the "Subject Characteristics" table- lots of boring summary statistics.

Vignette: https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html

```{r eval = FALSE}
library(tableone)
```

By default, its output is probably not what we want (though we can see hints of it down at the bottom)...

```{r}
CreateTableOne(data = mazes)
```

We need to tell it a bit about what we are looking for:

```{r}

## Vector of variables to summarize
my_num_vars <- c("ca", "viq", "content", "filler", "rep", "rev", "fs", "cued", "not_cued")

## Vector of categorical variables that need transformation
my_cat_vars <- c("dx", "activity")

## Create a TableOne object
tab2 <- CreateTableOne(vars = my_num_vars, data = mazes, factorVars = my_cat_vars)
print(tab2, showAllLevels = TRUE)
```

If we want the summary statistics to be computed on a per-strata basis, we can ask for that:

```{r}
# Another 
tab3 <- CreateTableOne(vars = my_num_vars, strata = "dx" , data = mazes)
tab3
```

## `table1`

Note that there is a similarly-named package called `table` that does something very similar, but has a different API and slightly different formatting capabilities:

```{r}
library(table1)

table1(~ ca + viq + content + filler + rep + rev+ fs+ cued + not_cued | dx, data=mazes) 

```

One nice thing about `table1` is that it makes more complex tables easier; for example, we can stratify by diagnosis _and_ ADOS activity as follows:

```{r}
table1(~ ca + viq + content + filler + rep + rev+ fs+ cued + not_cued | activity * dx, data=mazes) 

```

This is, of course, just for demonstration purposes! Please do not make a table this large. :-D

Note that `table1`'s output is always HTML, which may or may not be what you want. It has numerous options for customization and formatting, etc.

# The `DT` package

`DT` lets us produce interactive data tables for use in HTML reports that go far beyond `kable`'s capabilities.

An excellent tutorial on DT is available at https://rstudio.github.io/DT/.

"Out of the box", `datatable` gives us a very usable table:

```{r}
datatable(mazes)
```

There are numerous ways we can customized and configure things from here!

# Finally, fonts!

## Your friend, `extrafont`

The [`extrafont` package](https://github.com/wch/extrafont) is the best place to start with fonts in R and ggplot.


```{r}
library(extrafont)
# font_import() # build the font database; only need to call this once (or after installing new fonts)
loadfonts() # _load_ the font database; should call this once per session
```


**Note:** Make sure to follow _all_ installation instructions from `github`!

**Note:** Also, you may need to install a custom version of one of `extrafont`'s dependencies; if you get a bunch of errors about "No Font Name" and no fonts end up being imported, see this [discussion](https://github.com/wch/extrafont/issues/88).

You can access fonts on your local system in `ggplot` by using the `theme()` function to set the relevant properties of your figure.

Let's set up a basic scatterplot to start from:

```{r,echo=FALSE}
library(palmerpenguins)

demo_plt <- penguins %>% ggplot(aes(bill_length_mm, bill_depth_mm, color=species)) + 
  geom_point(size=3) +
  scale_color_discrete("Species") +
  labs(x="Bill Length (mm)", y="Bill Width (mm)", title="Palmer's Penguins") 
demo_plt
```

Out of the box, the default `ggplot` theme uses pretty boring fonts. How might we change that?

In `ggplot`, the `theme()` function lets us control the appearance of various aspects of the plot. There are [many, many properties](https://ggplot2.tidyverse.org/reference/theme.html) we can tweak; we will begin by setting the global `text` property for all text on the entire plot.

Let's try changing that font a bit:

```{r}
demo_plt + theme(text=element_text(family="Optima"))
```

```{r}
demo_plt + theme(text=element_text(family="Palatino"))
```

```{r}
demo_plt + theme(text=element_text(family="Zapfino"))
```

Note that the specific _name_ that you must use to refer to the font is not exactly obvious. It is not the font name, but rather the font's "family name". Because of some under-the-hood details about the way that fonts work, this is a bit different from what you might see in e.g. a font selection menu in Word. You can see a list of all the fonts that R knows about by using the `fonttable()` command:

```{r}
fonttable() %>% glimpse()
```

```{r}
fonttable() %>% head()
```


```{r}
fonttable() %>% select(FullName, FamilyName, FontName) %>% head(20)
```


```{r}
fonttable() %>% select(FamilyName) %>% distinct()
```


Long story short, you'll want to use the value in `FamilyName` to refer to your font of interest.

What if we wanted to have a different font for the title? The `plot.title` property is where we would want to look:

```{r}
demo_plt + theme(
  text=element_text(family="Palatino"),
  plot.title=element_text(family="Zapfino")
  )
```

We can follow this pattern to change anything we like about the text in different parts of the plot:

```{r}
demo_plt + theme(
  text=element_text(family="Palatino"),
  plot.title=element_text(family="Zapfino", size=8, color="blue"),
  axis.title.x = element_text(family="Optima", size=24)
  )
```


## Installing new fonts

This part is a bit beyond the scope of this class, but the upshot is that after you install new fonts in your computer, you should tell `extrafont` about it using the `font_import()` command.


## OTF Fonts:

`extrafont` only knows about TrueType fonts (ones whose file ends in `.ttf`); that's why `fonttable()`'s output is missing so many fonts on my system. 
If you have an OpenType font you want to use with R, your best bet probably involves the [`showtext` package](https://cran.rstudio.com/web/packages/showtext/vignettes/introduction.html).

You'll need to know the exact filename of the font you want to add; on a Mac, you can use the built-in FontBook application to help you find this:

```{r}
library(showtext)
font_add("Sofa Sans Regular", "FaceType - SofaSansHand-Regular.otf") # register the OTF font
```

Then, you just refer to it in your `element_text()` like any other font:

```{r}
demo_plt + theme(
  text=element_text(family="Sofa Sans")
)
```

## Using Google Fonts

`showtext` also has utility functions for working with Google Fonts. Google Fonts provides a wide variety of freely-usable font families, and is an amazing resource.

Here, we install and use the ["Oleo Script"](https://fonts.google.com/specimen/Oleo+Script) font.

```{r}
font_add_google("Oleo Script")
showtext_auto() # For Google Fonts specifically, showtext may need this
demo_plt + theme(
  text=element_text(family="Oleo Script")
)

```


## A Caveat and a Plea

It is important to note that both `extrafont` and `showtext` work best for _raster_ images; there are certain quirks that you are likely to encounter when producing PDF output using non-standard fonts (or, depending on the font, a plot that uses text containing text from the non-BMP regions of Unicode, like emojis). 
The story of "making PDFs with non-standard fonts in R" is extremely long and tedious; if you are actually interested, set up a meeting and I will be happy to go over it with you.

The upshot of it all is that if you want to make a PDF from `ggplot` using non-standard fonts or with non-BMP Unicode characters, your best bet is to use `showtext`, which _will_ produce a perfectly nice PDF file with whatever font and text you want... with one caveat: text selection (for search, copy-and-paste, etc.) will not be possible in the resulting PDFs. 
The file will look fine for everybody, it will print fine for everybody, it will scale fine to any resolution, and will do all the things a PDF should do, _other_ than allow text extraction.

If this is a dealbreaker for your particular situation, my honest advice is to find a different font to use- pick a TTF font (or figure out how to convert your OTF font to a TTF font), use `extrafont`, and move on with your life. 
Life is short and fleeting, the grave is cold and lonely, the climate is changing, and the full day that you will spend debugging this situation is a day of your life that you can never get back.
Nobody, on their deathbed, has ever wished that they had spent more time debugging mysterious error messages involving multiple obscure pieces of thirty-five-year-old software in order to get a PDF file to behave slightly better. 
Please, spend that day with your family, or go enjoy the coastline while it's still there, or do literally anything else _other_ than try and get R to produce proper PDF output with OTF fonts.

And note that this is advice coming from somebody who actually _enjoys_ dealing with bizarre font and Unicode issues, is a perfectionist when it comes to fonts and PDF files, and who eventually decided to acknowledge reality and just use `showtext`.

If you absolutely _must_ ignore my advice, here are two places to start:

- [Section 14.6](https://r-graphics.org/recipe-output-fonts-pdf) of the [R Graphics Cookbook](https://r-graphics.org)
- [My own page](https://www.bedrick.org/notes/fonts-and-ggplot-and-pdf/) on the subject, which includes some more detailed discussion and a workaround for what is going wrong with PDFs made using `extrafont` and non-standard TTF fonts
  - Note that this page will offer neither advice nor solace for those suffering from the OTF font situation
