---
title: "Lab 01b: BMI 5/625"
subtitle: "Introducing ggplot"
author: "Steven Bedrick"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
knit: pagedown::chrome_print
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300,
  fig.align = "center", out.width = "80%", cache = FALSE)
library(tidyverse)
library(palmerpenguins)

```

## Introducing `ggplot`

* Goal for this session: a very quick `ggplot` refresher

--

* Meet our dataset:

--

  * [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/)

--

```{r}
glimpse(penguins)
```



---

## Sidebar: Why not Fisher's Irises?

```{r, echo=FALSE, fig.height=4}
iris %>% ggplot(aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + geom_point()
```

--

* Hint: The original citation for that data is:

--

  * R. A. Fisher (1936) "The use of multiple measurements in taxonomic problems." _Annals of Eugenics_ 7(2): 179-188

--

* `r emo::ji("palm")`

---

## So: Penguins!

```{r}
glimpse(penguins)
```

---

## Core `ggplot` concepts

--

* _Aesthetics_ map data to visual properties of the plot

--

* _Geometries_ ("geoms") actually put "ink on the page"

--

  * Each type of graph (scatterplot, etc.) has a corresponding geom

--

  * Different geoms attend to different aesthetics
  
--

    * e.g., `geom_point` cares about `x` and `y`
--

  * Multiple geoms can be combined on the same plot

--

* _Scales_ control axes, fills, etc.

--

* _Themes_ control visual properties (fonts, background colors, etc.)

---
class: center, middle

## `ggplot` is an _opinionated_ tool!

---

## Data considerations

--

* By default, `ggplot` assumes "tidy" data

--

* Think: one "row" per mark on the graph...

--

* ... and then the various properties for each mark in columns.

---

## Our first plot

```{r, eval=FALSE}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, y=bill_depth_mm)
  ) + geom_point()
```
--
```{r, echo=FALSE, fig.height=4}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, y=bill_depth_mm)
  ) + geom_point()
```

---

## What about color?


```{r, eval=FALSE}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, y=bill_depth_mm, color=species)
  ) + geom_point()
```
--
```{r, echo=FALSE, fig.height=4}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, y=bill_depth_mm, color=species)
  ) + geom_point()
```

---

## Want a different plot? Try a different geom

```{r, fig.height=4}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, fill=species)
  ) + geom_histogram()
```

---

## We can specify aesthetic values by hand:

```{r, fig.height=4}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, fill=species)
  ) + geom_histogram(alpha=0.6)
```

---

## Customizing other aspects of the plot

```{r, fig.height=4}
penguins %>% ggplot(
  mapping=aes(x=bill_length_mm, fill=species)
  ) + geom_histogram(alpha=0.6) +
  labs(x="Bill Length (mm)", y="Frequency") +
  ggtitle("Bill Length, by species")
```

