---
title: "Lab 04: BMI 5/625"
subtitle: "Working with Tidy Data"
author: "Alison Hill (with modifications by Steven Bedrick)"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
knit: pagedown::chrome_print
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300,
  fig.align = "center", out.width = "80%", cache = FALSE)
library(tidyverse)
library(here)
```


class: middle, center, inverse

# ⌛️

## Let's review 

---

## Data wrangling to date! 


.pull-left[

From `dplyr`:
- `filter`
- `arrange`
- `mutate`
- `group_by`
- `summarize`
- `glimpse`
- `distinct`
- `count`
- `tally`
- `pull`
- `top_n`
- `case_when`

]

--

.pull-right[
Let's add from `dplyr`:

- `select`

From `tidyr`:

- `pivot_longer`
- `pivot_wider`

Plus 1 other package:

- `skimr::skim`

]

---
# The Great British Baking Data Set

```{r echo=FALSE}
cakes_raw <- read_csv(here::here("data", "tidy-gbbo-cakes.csv"))
glimpse(cakes_raw)
```

---

# Un-tidy cakes

```{r echo = FALSE}
cakes_untidy <- read_csv(here::here("data", "tidy-gbbo-cakes.csv"), col_types = cols(series = col_factor(levels = NULL))) %>% 
  count(series, challenge, cake) %>% 
  drop_na(cake) %>% 
  rename(bakes = n) %>% 
  spread(cake, bakes, fill = 0) 

cakes_split <- cakes_untidy %>% 
  split(.$series) 
```


.pull-left[
```{r echo = FALSE}
cakes_split[[1]]
cakes_split[[2]]
```
]

--


.pull-right[
```{r echo = FALSE}
cakes_split[[3]]
cakes_split[[4]]
```
]

--

Four seasons, four datasets...

--

Each row: a challenge type ("signature" vs. "showstopper") and a count of entries by type

---

# Still un-tidy cakes

.pull-left[

```{r}
cakes_untidy %>% 
  bind_rows()
```
]

--

.pull-right[
At least now it's a single dataframe...
]
---

# Finally tidy cakes

```{r}
cakes_tidy <- cakes_untidy %>% 
  pivot_longer(cake:pie_tart, 
               names_to="bake_type", 
               values_to="num_bakes") %>% 
  arrange(series)
cakes_tidy
```


---

# What about changing types?

```{r}
cakes_tidy <- cakes_untidy %>% 
  pivot_longer(cake:pie_tart, 
               names_to="bake_type", 
               names_transform = list(bake_type=as.factor),
               values_to="num_bakes") %>% 
  arrange(series)
cakes_tidy
```

---
class: middle, inverse, center

## Know Your Tidy Data

---

```{r}
glimpse(cakes_tidy)
```


---

```{r,eval=FALSE}
library(skimr)
skim(cakes_tidy)
```

```{r,echo=FALSE}
library(skimr)
cakes_skim <- skim(cakes_tidy)
```
--

```{r, echo=FALSE}
summary(cakes_skim)
```

---

```{r,eval=FALSE}
library(skimr)
skim(cakes_tidy)
```

--

```{r,echo=FALSE}
cakes_skim %>% yank("character")
```

--

```{r,echo=FALSE}
cakes_skim %>% yank("factor")
```
--

```{r,echo=FALSE}
cakes_skim %>% yank("numeric")
```


---

```{r}
skim(cakes_tidy) %>% 
  summary() 
```


---
class: middle, inverse, center

## Benefits of Tidy Data


---


```{r}
cakes_tidy %>% 
  count(challenge, bake_type, wt = num_bakes, sort = TRUE)
```

---


```{r}
cakes_tidy %>% 
  count(series, bake_type, wt = num_bakes)
```

---

```{r}
library(skimr)

cakes_tidy %>% 
  group_by(bake_type) %>% 
  select_if(is.numeric) %>% 
  skim() %>% summary
```

See: https://suzanbaert.netlify.com/2018/01/dplyr-tutorial-1/

---


```{r}
cakes_by_series <- cakes_tidy %>% 
  count(series, bake_type, wt = num_bakes) 
cakes_by_series
```

---


```{r fig.height = 3, fig.width = 4}
ggplot(cakes_by_series, aes(x = series, y = n, 
                            color = bake_type,
                            group = bake_type)) +
  geom_point() +
  geom_line() +
  expand_limits(y = 0)
```

---
class: middle, inverse, center

## `select()`: Your new best friend

---

# Selection Helpers

`dplyr` gives us helpful syntax for selecting columns:

```{r}
cakes_raw %>% head(4)
```

What if we only want _some_ of the columns?

---

# `dplyr::select()` to the rescue

```{r}
cakes_raw %>% select(cake)
```


---

# `dplyr::select()` to the rescue

```{r}
cakes_raw %>% select(cake, baker) %>% head(4)
```

But this is only the beginning!

---

# ... All columns _other_ than `cake`

```{r}
cakes_raw %>% select(!cake) %>% head(4)
```


---

# Columns that _start_ with a string?

```{r}
cakes_raw %>% select(starts_with("c"))
```

---

# The last column...

```{r}
cakes_raw %>% select(last_col()) %>% head(4)
```


---

# A _range_ of contiguous columns

```{r}
cakes_raw %>% select(baker:cake) %>% head(4)
```

---

# There are many other helpers:

Matching columns by name:
* `starts_with()`/`ends_with()`
* `contains()`
* `num_range()` (for matching numerical ranges: think columns named for years, quarters, etc.)

See the `select` help page for more examples...

---

# Many Tidyverse functions work with select helpers

```{r, echo=FALSE}
#install.packages("billboard")
library(billboard)
```


```{r}
billboard %>% glimpse
```

---
# Many Tidyverse functions work with select helpers

```{r}
billboard %>% 
  pivot_longer(cols=starts_with("wk"),
               names_to = "week",
               names_prefix = "wk",
               values_to = "rank"
  ) %>% head(10)
```
---
class: middle, inverse, center

## `janitor`: Your _other_ new best friend

---

Often, data comes to us in... a less than pristine state:

--

```{r, echo=FALSE}
library(readxl)
library(here)
gapmnd <- read_excel(here("data/gapminder_broadband_per_100.xlsx"))
```

```{r}
glimpse(gapmnd)
```

--

Note the very inconvenient column names...

---

## The `janitor` package is here to help!

```{r}
gapmnd %>% janitor::clean_names() %>% glimpse()
```

---

### `janitor` has many other capabilities...

--

* Transforming columns

--

* Removing empty rows/columns

--

* Collapsing sets of values values to `NA`, as needed

--

It also has a very nice cross-tabulation syntax (`tabyl()`)!


---


## You have 2 _challenges_ today!

Described [here](04-challenge_rev.html)

Also see a reference walkthrough [here](04-distributions.html)


---
class: middle, inverse, center

# 🍱

## Tidy Data:

http://r4ds.had.co.nz/tidy-data.html

http://moderndive.com/4-tidy.html

http://vita.had.co.nz/papers/tidy-data.html

https://github.com/jennybc/lotr-tidy#readme